---
title: "Differential Expression & GSEA Enrichment Plots"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
    self_contained: true
params:
  de_res_fn:  "/Users/hiteshkore/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Hitesh_parker_lab/github/downstream_transcriptome_analysis/RNASeqWorkbench/Test_data/Shoaib/IC30-41PRO_IC7-41PRO_gsea.xlsx"
  method:     "deseq"
  outdir:     "/Users/hiteshkore/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Hitesh_parker_lab/github/downstream_transcriptome_analysis/RNASeqWorkbench/Test_data/Shoaib/Final_results/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,   # hide code from the rendered report
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 7
)

library(readxl)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

# Overview

This report generates:

1. A **volcano plot** of differential expression (DE) results.
2. **GSEA/GO enrichment plots** (lollipop or bar style) for Biological
   Process (BP), Molecular Function (MF), Cellular Component (CC), and
   MSigDb Hallmark / Immune Signature gene sets.

Every plot is rendered inline below in this single HTML report, and also
saved to `outdir` as an individual **PDF** file.

```{r params}
# Pull knit parameters into plain variables for readability below.
DE_res_fn  <- params$de_res_fn
method     <- params$method
outdir     <- params$outdir

# Make sure the output directory exists before we try to save anything into it.
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
```

```{r fn-check-input-columns}
#' Validate that a data frame has the columns required for a given method.
#'
#' @param df     Data frame to check.
#' @param method Character. Currently supports "deseq".
#' @return TRUE (invisibly) if valid; otherwise throws an error listing
#'         which required columns are missing.
check_input_columns <- function(df, method) {

  if (method == "deseq") {
    required_columns <- c(
      "Gene", "baseMean", "log2FoldChange", "lfcSE",
      "pvalue", "padj", "regulation", "gene_name"
    )

    missing_columns <- setdiff(required_columns, colnames(df))

    if (length(missing_columns) > 0) {
      stop(
        "For method = 'deseq', the following columns are missing: ",
        paste(missing_columns, collapse = ", ")
      )
    }
  }

  invisible(TRUE)
}

#' Validate that a data frame contains a set of arbitrary column names.
#'
#' Used by the enrichment-plot functions so that a wrong/missing column
#' name (e.g. a typo in `ratio_col`) fails fast with a clear message,
#' instead of erroring deep inside ggplot2's internals.
#'
#' @param df            Data frame to check.
#' @param required_cols Character vector of column names to look for.
#'                       NULL entries (e.g. an unset optional argument)
#'                       are ignored.
validate_columns <- function(df, required_cols) {
  required_cols <- required_cols[!sapply(required_cols, is.null)]

  missing_cols <- setdiff(required_cols, colnames(df))

  if (length(missing_cols) > 0) {
    stop(
      "The following column(s) were not found in the data frame: ",
      paste(missing_cols, collapse = ", "),
      "\nAvailable columns are: ",
      paste(colnames(df), collapse = ", "),
      call. = FALSE
    )
  }

  invisible(TRUE)
}

#' Read one sheet from the results workbook, failing with a clear message
#' (listing the sheets that DO exist) if the requested sheet name isn't found.
#'
#' @param path       Path to the .xlsx file.
#' @param sheet_name Name of the sheet to read.
#' @return A data frame (tibble) with that sheet's contents.
read_sheet_safely <- function(path, sheet_name) {
  available_sheets <- excel_sheets(path)

  if (!sheet_name %in% available_sheets) {
    stop(
      "Sheet '", sheet_name, "' not found in ", path,
      ".\nAvailable sheets are: ", paste(available_sheets, collapse = ", ")
    )
  }

  read_excel(path, sheet = sheet_name)
}
```

```{r fn-volcano-plot}
#' Build a publication-style volcano plot.
#'
#' @param features          Character vector of feature/gene identifiers.
#' @param pvalue            Numeric vector of (adjusted) p-values.
#' @param fold_change       Numeric vector of log2 fold changes.
#' @param selected_features Character vector of features to label on the
#'                           plot (only labeled if also significant).
#' @param lfc_cutoff        |log2FC| threshold used to call up/down (default 1).
#' @param p_cutoff          p-value significance threshold (default 0.05).
#' @param title             Plot title.
#' @param x_breaks_params   Optional c(lower, upper, increment) for the x-axis.
#'                           If NULL, breaks are computed automatically.
#' @param y_breaks_params   Optional c(lower, upper, increment) for the y-axis.
#'                           If NULL, breaks are computed automatically.
#' @return A ggplot object.
plot_volcano_plot <- function(features, pvalue, fold_change,
                               selected_features = NULL,
                               lfc_cutoff = 1, p_cutoff = 0.05,
                               title = "Volcano Plot of Differential Expression",
                               x_breaks_params = NULL,
                               y_breaks_params = NULL) {

  DE_data <- data.frame(features, pvalue, fold_change) %>% na.omit()

  DE_data$log_p <- -log10(DE_data$pvalue)

  # Classify each feature as up/down/unchanged using the fold-change and
  # p-value cutoffs supplied above.
  DE_data$significance <- "Unchanged"
  DE_data$significance[DE_data$fold_change >  lfc_cutoff & DE_data$pvalue < p_cutoff] <- "Upregulated"
  DE_data$significance[DE_data$fold_change < -lfc_cutoff & DE_data$pvalue < p_cutoff] <- "Downregulated"
  DE_data$significance <- factor(DE_data$significance,
                                  levels = c("Downregulated", "Unchanged", "Upregulated"))

  # Only label features that are both in the user-selected list AND significant,
  # and nudge labels left/right depending on which side of zero they sit.
  DE_data <- DE_data %>%
    mutate(
      label   = ifelse(features %in% selected_features & significance != "Unchanged",
                        features, NA),
      nudge_x = ifelse(fold_change > 0, lfc_cutoff, -lfc_cutoff)
    )

  n_up   <- sum(DE_data$significance == "Upregulated")
  n_down <- sum(DE_data$significance == "Downregulated")

  x_max <- max(abs(DE_data$fold_change), na.rm = TRUE) * 1.1
  y_max <- max(DE_data$log_p, na.rm = TRUE) * 1.1

  # Axis breaks: use user-supplied c(lower, upper, increment) if given,
  # otherwise fall back to sensible automatic breaks.
  if (!is.null(x_breaks_params)) {
    x_breaks <- seq(x_breaks_params[1], x_breaks_params[2], by = x_breaks_params[3])
    x_lim    <- c(x_breaks_params[1], x_breaks_params[2])
  } else {
    x_breaks <- seq(floor(-x_max / 2) * 2, ceiling(x_max / 2) * 2, by = 2)
    x_lim    <- c(-x_max, x_max)
  }

  if (!is.null(y_breaks_params)) {
    y_breaks <- seq(y_breaks_params[1], y_breaks_params[2], by = y_breaks_params[3])
    y_lim    <- c(y_breaks_params[1], y_breaks_params[2])
  } else {
    y_breaks <- seq(0, ceiling(y_max / 3) * 3, by = 3)
    y_lim    <- c(0, y_max)
  }

  ggplot(DE_data, aes(x = fold_change, y = log_p, color = significance)) +
    geom_point(alpha = 0.6, size = 1.8, shape = 16) +
    scale_color_manual(
      values = c("Upregulated" = "#D62728",
                 "Downregulated" = "#1F77B4",
                 "Unchanged" = "grey75"),
      drop = FALSE
    ) +
    geom_hline(yintercept = -log10(p_cutoff), linetype = "dashed", color = "grey40", linewidth = 0.4) +
    geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff), linetype = "dashed", color = "grey40", linewidth = 0.4) +
    geom_text_repel(
      aes(label = label),
      nudge_x = DE_data$nudge_x,
      direction = "y",
      size = 3,
      color = "black",
      fontface = "italic",
      box.padding = 0.4,
      point.padding = 0.3,
      segment.color = "grey30",
      segment.size = 0.3,
      max.overlaps = Inf,
      min.segment.length = 0
    ) +
    annotate("text", x = x_lim[1] * 0.95, y = y_lim[2], label = paste0("Down: ", n_down),
             hjust = 0, size = 3.5, color = "#1F77B4") +
    annotate("text", x = x_lim[2] * 0.95, y = y_lim[2], label = paste0("Up: ", n_up),
             hjust = 1, size = 3.5, color = "#D62728") +
    labs(
      title = title,
      x = expression(log[2] ~ "(Fold Change)"),
      y = expression(-log[10] ~ "(P value)")
    ) +
    coord_cartesian(xlim = x_lim, ylim = y_lim) +
    scale_x_continuous(breaks = x_breaks) +
    scale_y_continuous(breaks = y_breaks) +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.title = element_text(face = "bold")
    )
}
```

```{r fn-go-plots}

#' Generate one enrichment plot for a single results tab.
#' @param de_res_fn  Path to the results workbook.
#' @param tab        Sheet name to read (e.g. "BP", "KEGG", "mSigDb_HM").
#' @param plot_type  "lollipop" or "bar" for this specific tab.
#' @param title      Plot title.
#' @param outdir     Directory to save the PDF into.
#' @param width      PDF export width, in inches.
#' @param height     PDF export height, in inches.
#' @param ...        Passed through to plot_go_enrichment() (term_col, ratio_col,
#'                   pval_col, count_col, color_direction, show_count_label).
#' @return A ggplot object.
generate_enrichment_plot <- function(de_res_fn, tab, plot_type, title, outdir,
                                      width = 9, height = 7, ...) {

  df <- read_sheet_safely(de_res_fn, tab)

  filtered_df <- df %>% filter(include_in_plot == TRUE)

  p <- plot_go_enrichment(
    filtered_df,
    plot_type = plot_type,
    title     = title,
    ...
  )

  export_plot(p, paste0("go_", tolower(tab), "_enrichment"), outdir,
              width = width, height = height)

  p
}
# small helper used below: `a %||% b` returns a unless it's NULL, else b
`%||%` <- function(a, b) if (is.null(a)) b else a

#' Read the plot_instructions sheet and generate every enrichment plot it lists,
#' using each row's own plot type, title, and PDF dimensions.
#'
#' Expected columns in the instructions sheet: Tab, Plot_type, Title,
#' Plot_width, Plot_height. Column names are trimmed of whitespace and
#' matched case-insensitively, so "Tittle " or "PLOT_TYPE" etc. still work.
#'
#' @param de_res_fn          Path to the results workbook.
#' @param outdir             Directory to save PDFs into.
#' @param instructions_sheet Name of the sheet mapping Tab -> plot settings
#'                           (default "plot_instructions").
#' @param ...                Passed through to plot_go_enrichment() (term_col,
#'                           ratio_col, pval_col, count_col, color_direction,
#'                           show_count_label).
#' @return A named list of ggplot objects, one per tab (names = tab names).
generate_all_enrichment_plots <- function(de_res_fn, outdir,
                                           instructions_sheet = "plot_instructions",
                                           ...) {

  instructions <- read_sheet_safely(de_res_fn, instructions_sheet)

  # Normalise column names: trim whitespace, so a header like "Tittle "
  # (with a trailing space) is matched reliably.
  colnames(instructions) <- trimws(colnames(instructions))

  # Map expected logical names -> whatever the actual (trimmed) column is,
  # matched case-insensitively so "Title"/"TITLE"/"Tittle" all resolve.
  find_col <- function(df, target) {
    match_idx <- which(tolower(colnames(df)) == tolower(target))
    if (length(match_idx) == 0) return(NULL)
    colnames(df)[match_idx[1]]
  }

  tab_col    <- find_col(instructions, "Tab")
  type_col   <- find_col(instructions, "Plot_type")
  title_col  <- find_col(instructions, "Title")
  width_col  <- find_col(instructions, "Plot_width")
  height_col <- find_col(instructions, "Plot_height")

  required <- c(tab = tab_col, plot_type = type_col, title = title_col,
                width = width_col, height = height_col)

  if (any(sapply(required, is.null))) {
    stop(
      "The following required column(s) are missing from '", instructions_sheet,
      "': ", paste(names(required)[sapply(required, is.null)], collapse = ", "),
      "\nAvailable columns are: ", paste(colnames(instructions), collapse = ", ")
    )
  }

  plots <- setNames(vector("list", nrow(instructions)), instructions[[tab_col]])

  for (i in seq_len(nrow(instructions))) {
    tab    <- instructions[[tab_col]][i]
    ptype  <- tolower(trimws(instructions[[type_col]][i]))
    title  <- instructions[[title_col]][i]
    width  <- as.numeric(instructions[[width_col]][i])
    height <- as.numeric(instructions[[height_col]][i])

    plots[[tab]] <- generate_enrichment_plot(
      de_res_fn = de_res_fn,
      tab       = tab,
      plot_type = ptype,
      title     = title,
      outdir    = outdir,
      width     = width,
      height    = height,
      ...
    )
  }

  plots
}



#' Prepare a GSEA/GO results data frame for plotting.
#'
#' Handles two shapes of "ratio" column: a GeneRatio-style "5/120" string
#' (as in enrichGO results), or a plain numeric column (e.g. NES from
#' gseGO/gseKEGG results). Also derives a gene Count from `core_enrichment`
#' when no explicit count column is supplied.
#'
#' @param df        Enrichment results data frame.
#' @param ratio_col Name of the column to use for the x-axis (default "NES").
#' @param count_col Name of the column to size points / label bars by.
#'                   If NULL, Count is derived from `core_enrichment`.
#' @return The input data frame with `GeneRatio_num` and `Count` columns added.
prepare_enrichment_df <- function(df, ratio_col, count_col) {

  plot_df <- df

  if (is.character(plot_df[[ratio_col]])) {
    # e.g. "5/120" -> 0.0417
    parts <- strsplit(as.character(plot_df[[ratio_col]]), "/")
    plot_df$GeneRatio_num <- sapply(parts, function(x) as.numeric(x[1]) / as.numeric(x[2]))
  } else {
    plot_df$GeneRatio_num <- plot_df[[ratio_col]]
  }

  if (is.null(count_col)) {
    plot_df$Count <- sapply(strsplit(as.character(plot_df$core_enrichment), "/"), length)
  }

  plot_df
}

#' Lollipop-style enrichment plot (points on stems, ordered by GeneRatio/NES).
#'
#' @param df               Enrichment results data frame.
#' @param term_col         Column holding the term description (y-axis labels).
#' @param ratio_col        Column used for the x-axis (default "NES").
#' @param pval_col         Column used to color points (default "p.adjust").
#' @param count_col        Column used to size points. NULL derives it from
#'                         `core_enrichment`.
#' @param title            Plot title.
#' @param color_direction  Direction of the viridis color scale (-1 or 1).
#' @return A ggplot object.
plot_go_lollipop <- function(df, term_col = "Description", ratio_col = "NES",
                              pval_col = "p.adjust", count_col = NULL,
                              title = "GO Enrichment", color_direction = -1) {

  validate_columns(df, c(term_col, ratio_col, pval_col))

  plot_df <- prepare_enrichment_df(df, ratio_col, count_col)
  if (is.null(count_col)) count_col <- "Count"

  # Order terms by GeneRatio/NES so the plot reads top-to-bottom, largest first.
  plot_df <- plot_df %>% arrange(desc(GeneRatio_num))
  plot_df[[term_col]] <- factor(
    plot_df[[term_col]],
    levels = rev(plot_df[[term_col]][order(plot_df$GeneRatio_num)])
  )

  ggplot(plot_df, aes(x = GeneRatio_num, y = .data[[term_col]])) +
    geom_vline(xintercept = 0, color = "grey40", linewidth = 0.6) +
    geom_segment(aes(x = 0, xend = GeneRatio_num,
                      y = .data[[term_col]], yend = .data[[term_col]]),
                 color = "black", linewidth = 0.5) +
    geom_point(aes(color = .data[[pval_col]], size = .data[[count_col]])) +
    scale_color_viridis_c(name = "p.adjust", direction = color_direction) +
    scale_size_continuous(name = "Count", range = c(2, 8)) +
    labs(title = title, x = "Gene Ratio", y = NULL) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.title  = element_text(hjust = 0.5, size = 16),
      axis.text.y = element_text(color = "grey30"),
      axis.line.x = element_line(color = "black", linewidth = 0.5),
      axis.ticks.x = element_line(color = "black")
    )
}

#' Bar-style enrichment plot (bars colored by p.adjust, ordered by GeneRatio/NES).
#'
#' @inheritParams plot_go_lollipop
#' @param show_count_label Logical; if TRUE, annotate each bar with its Count.
#' @return A ggplot object.
plot_go_barplot <- function(df, term_col = "Description", ratio_col = "NES",
                             pval_col = "p.adjust", count_col = NULL,
                             title = "GO Enrichment", color_direction = -1,
                             show_count_label = TRUE) {

  validate_columns(df, c(term_col, ratio_col, pval_col))

  plot_df <- prepare_enrichment_df(df, ratio_col, count_col)
  if (is.null(count_col)) count_col <- "Count"

  plot_df <- plot_df %>% arrange(desc(GeneRatio_num))
  plot_df[[term_col]] <- factor(
    plot_df[[term_col]],
    levels = rev(plot_df[[term_col]][order(plot_df$GeneRatio_num)])
  )

  p <- ggplot(plot_df, aes(x = GeneRatio_num, y = .data[[term_col]])) +
    geom_vline(xintercept = 0, color = "grey40", linewidth = 0.6) +
    geom_col(aes(fill = .data[[pval_col]]), width = 0.7, color = "black", linewidth = 0.3) +
    scale_fill_viridis_c(name = "p.adjust", direction = color_direction) +
    labs(title = title, x = "Gene Ratio", y = NULL) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.title  = element_text(hjust = 0.5, size = 16),
      axis.text.y = element_text(color = "grey30"),
      axis.line.x = element_line(color = "black", linewidth = 0.5),
      axis.ticks.x = element_line(color = "black")
    )

  if (show_count_label) {
    p <- p + geom_text(
      aes(label = .data[[count_col]],
          hjust = ifelse(GeneRatio_num >= 0, -0.2, 1.2)),
      size = 3, color = "black"
    )
  }

  p
}

#' Dispatch wrapper: build either a lollipop or bar enrichment plot.
#'
#' @inheritParams plot_go_lollipop
#' @param plot_type       One of "lollipop" or "bar".
#' @param show_count_label Passed through to plot_go_barplot() (ignored for lollipop).
#' @return A ggplot object.
plot_go_enrichment <- function(df,
                                plot_type  = c("lollipop", "bar"),
                                term_col   = "Description",
                                ratio_col  = "NES",
                                pval_col   = "p.adjust",
                                count_col  = NULL,
                                title      = "GO Enrichment",
                                color_direction   = -1,
                                show_count_label  = TRUE) {

  # match.arg() validates plot_type against the allowed options and gives
  # a clear error message if an unsupported value (or a typo) is passed.
  plot_type <- match.arg(plot_type)

  if (plot_type == "lollipop") {
    plot_go_lollipop(
      df = df, term_col = term_col, ratio_col = ratio_col,
      pval_col = pval_col, count_col = count_col,
      title = title, color_direction = color_direction
    )
  } else {
    plot_go_barplot(
      df = df, term_col = term_col, ratio_col = ratio_col,
      pval_col = pval_col, count_col = count_col,
      title = title, color_direction = color_direction,
      show_count_label = show_count_label
    )
  }
}
```

```{r fn-export-plot}
#' Save a ggplot as a static PDF file in `outdir`. No other files are created.
#'
#' @param plot     A ggplot object.
#' @param filename Base filename (no extension), e.g. "go_bp_enrichment".
#' @param outdir   Directory to write the PDF into.
#' @param width,height Dimensions in inches.
export_plot <- function(plot, filename, outdir, width = 9, height = 7) {
  pdf_path <- file.path(outdir, paste0(filename, ".pdf"))
  ggsave(pdf_path, plot = plot, width = width, height = height)
  invisible(pdf_path)
}
```

---

# Load differential expression results

```{r load-de-data}
DE_data <- read_sheet_safely(DE_res_fn, "DE_genes")

# Fail fast with a clear message if expected columns are missing.
check_input_columns(DE_data, method)
```

# Volcano plot

```{r volcano-plot, fig.cap="Volcano plot of differential expression results."}
highlight_genes <- unique(DE_data$highlight[DE_data$highlight != "None"])

volcano_plot <- plot_volcano_plot(
  features          = DE_data$gene_name,
  pvalue            = DE_data$padj,
  fold_change       = DE_data$log2FoldChange,
  selected_features = highlight_genes,
  x_breaks_params   = c(-9, 18, 3),
  y_breaks_params   = c(0, 35, 5)
)

volcano_plot

export_plot(volcano_plot, "volcano_plot", outdir)
```

# GO / GSEA enrichment plots

Generates enrichment plots for biological process, molecular functions, cellular components, hallmarks and immune signatures.

```{r go-bp, fig.cap="Enriched Biological Processes."}
all_plots <- generate_all_enrichment_plots(DE_res_fn, outdir)

for (tab in names(all_plots)) {
  cat("\n\n## ", tab, "\n\n")
  print(all_plots[[tab]])
  cat("\n\n")
}

```


